---
title: "DS311 - R Lab Assignment"
author: "Sara Noora Sundwell"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Assignment 1

* In this assignment, we are going to apply some of the build in data set in R for descriptive statistics analysis.
* To earn full grade in this assignment, students need to complete the coding tasks for each question to get the result.
* After finished all the questions, knit the document into HTML format for submission.

### Question 1

Using the **mtcars** data set in R, please answer the following questions.

```{r}
# Loading the data
data(mtcars)

# Head of the data set
head(mtcars)
```

a. Report the number of variables and observations in the data set.

```{r}
# Enter your code here!

Variables <- ncol(mtcars)
Observations <- nrow(mtcars)

# Answer:
print(paste("There are total of", Variables, "variables and", Observations, "observations in this data set."))
```



b. Print the summary statistics of the data set and report how many discrete and continuous variables are in the data set.

```{r}
# Enter your code here!

summary(mtcars)
str(mtcars)


# Answer:
print(paste("There are", ncol(mtcars) - sum(sapply(mtcars, is.numeric)), "discrete variables and", sum(sapply(mtcars, is.numeric)), "continuous variables in this data set."))
```



c. Calculate the mean, variance, and standard deviation for the variable **mpg** and assign them into variable names m, v, and s. Report the results in the print statement.

```{r}
# Enter your code here!

m <- mean(mtcars$mpg, na.rm = TRUE)
v <- var(mtcars$mpg, na.rm = TRUE)
s <- sd(mtcars$mpg, na.rm = TRUE)


# Answer
print(paste("The average of Mile Per Gallon from this data set is ", m , " with variance ", v , " and standard deviation", s , "."))
```



d. Create two tables to summarize 1) average mpg for each cylinder class and 2) the standard deviation of mpg for each gear class.

```{r}
# Enter your code here!

#Table 1

average_mpg_cylinder <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)


#Table 2

std_dev_mpg_gear <- aggregate(mpg ~ gear, data = mtcars, FUN = sd)

#Print tables
print("Average MPG for each Cylinder Class:")
print(average_mpg_cylinder)

print("Standard Deviation of MPG for each Gear Class:")
print(std_dev_mpg_gear)

```



e. Create a crosstab that shows the number of observations belong to each cylinder and gear class combinations. The table should show how many observations given the car has 4 cylinders with 3 gears, 4 cylinders with 4 gears, etc. Report which combination is recorded in this data set and how many observations for this type of car.

```{r}
# Enter your code here!

table <- table(mtcars$cyl, mtcars$gear)

# Print the crosstab
print(table)

# Find the most common car type
max_count <- max(table)
type <- which(table == max_count, arr.ind = TRUE)

print(paste("The most common car type in this data set is car with", type[1], "cylinders and", type[2], "gears. There are total of", max_count, "cars belong to this specification in the data set."))
```


*** 

### Question 2

Use different visualization tools to summarize the data sets in this question.

a. Using the **PlantGrowth** data set, visualize and compare the weight of the plant in the three separated group. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your findings.

```{r}
# Load the data set
data("PlantGrowth")

# Head of the data set
head(PlantGrowth)

# Enter your code here!

boxplot(weight ~ group, data = PlantGrowth,
        main = "Plant Growth by Group",
        xlab = "Group", ylab = "Weight",
        names = c("Control", "Treatment 1", "Treatment 2"))

# The boxplot shows that:
#         Group 1 (Control) exhibits the highest median plant weight, followed by Group 2 (Treatment 1)           and Group 3 (Treatment 2)

#         The variation in weight appears to be greatest in Group 1, while Groups 2 and 3 show less              variability.
#
```



b. Using the **mtcars** data set, plot the histogram for the column **mpg** with 10 breaks. Give labels to the title, x-axis, and y-axis on the graph. Report the most observed mpg class from the data set.

```{r}

# Print histogram
hist(mtcars$mpg, breaks = 10,
     main = "Histogram of Miles Per Gallon",
     xlab = "Miles Per Gallon", ylab = "Frequency")

# Observations
most_observed <- as.numeric(names(sort(table(mtcars$mpg), decreasing = TRUE)[1]))

# Answer
print(paste("Most of the cars in this data set are in the class of", most_observed, "mile per gallon."))
```



c. Using the **USArrests** data set, create a pairs plot to display the correlations between the variables in the data set. Plot the scatter plot with **Murder** and **Assault**. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your results from both plots.

```{r}
# Load the data set
data("USArrests")

# Head of the data set
head(USArrests)

# Enter your code here!
pairs(USArrests)
plot(USArrests$Murder, USArrests$Assault,
     main = "Murder vs. Assault",
     xlab = "Murder Rate", ylab = "Assault Rate")

# My findings
#   The plots show a strong link between murder and assault rates in US states.
#   States with high murder rates often have high assault rates too.
```


***

### Question 3

Download the housing data set from www.jaredlander.com and find out what explains the housing prices in New York City.

Note: Check your working directory to make sure that you can download the data into the data folder. 

```{r, echo=FALSE}
# Load and clean the housing data set 
download.file(url='https://www.jaredlander.com/data/housing.csv',
              destfile='data/housing.csv', mode='wb')
housingData <- read.csv('data/housing.csv')
housingData <- subset(housingData, 
                      select = c("Neighborhood", "Market.Value.per.SqFt", "Boro", "Year.Built"))
housingData <- na.omit(housingData)                   
```

a. Create your own descriptive statistics and aggregation tables to summarize the data set and find any meaningful results between different variables in the data set.

```{r}
# Head of the cleaned data set
head(housingData)

# Enter your code here!
summary(housingData)

# Average price per square foot by borough
avg_price_by_boro <- aggregate(Market.Value.per.SqFt ~ Boro, data = housingData, FUN = mean)
print(avg_price_by_boro)

# Correlation between Year Built and Market Value per SqFt
cor(housingData$Year.Built, housingData$Market.Value.per.SqFt)

# Boxplot of Market Value per SqFt by Borough
boxplot(Market.Value.per.SqFt ~ Boro, data = housingData, main = "Market Value per SqFt by Borough")

```

b. Create multiple plots to demonstrates the correlations between different variables. Remember to label all axes and give title to each graph.

```{r}
# Enter your code here!
plot(housingData$Year.Built, housingData$Market.Value.per.SqFt,
     xlab = "Year Built", ylab = "Market Value per SqFt",
     main = "Market Value vs. Year Built")

# Boxplot of market value by borough
boxplot(Market.Value.per.SqFt ~ Boro, data = housingData,
        main = "Market Value by Borough")

```

c. Write a summary about your findings from this exercise.

=> Enter your answer here!

So, what did we learn about NYC housing?
- New buildings cost more: Newer buildings are generally more expensive than older ones.
- Manhattan is the most expensive: If you want to live in Manhattan, be prepared to pay a lot of money.
- Other boroughs are cheaper: The other boroughs in NYC are generally more affordable than Manhattan.

Basically, if you want a fancy, new apartment in Manhattan, get ready to break the bank

